---
title: "final_project_DataCleaning"
author: "Ruth"
date: "2025-05-04"
output: html_document
---

importing my .csv file from johns Hopkins CSSE COVID-19 Data
```{r}
# Step 1: Install readr (if you haven't already)
install.packages("readr")

# Step 2: Load the readr package
library(readr)

# Step 3: Then run your read_csv line
data <- read_csv("owid-covid-data.csv")
install.packages("dplyr")   # Run this only once
library(dplyr) 
```


taking a closer look 
  - the dimension 
  - N/a values on each row/column 
  - duplicates 
  - negative values or weird values that dont fit int(outliers)
```{r}
#checking my columns 
colnames(data)
# looking at the unique values of each col. lorddd it is a lotttt
for (col in names(data)) {
  cat(" Column:", col, "\n")
  print(unique(data[[col]]))
  cat("\n--------------------------\n")
  
}
```

Here I am trying to look at the NA's in each column and display it nice and check for how may percentage of my data is null vals..
  - the data is sparse for most of my variables
  - I decided to gruop my variables based on what they are trying to measure 
    1 Geographic Identity columns
    2 Covid 19 cases data
    3 Covid 19 deaths data
    4 testing Data
    5 Vaccination Data
    6 Hospitilization and ICU data
    7 Goverment response 
    8 Demographics & Socioeconomic Indicator
    9 Excess Mortality

Just taking a quick look I most definetly don't need Hospitilization and ICU data, because it has >90% NA and it is not relevant for the scope of my project.. so I will drop these variables

```{r}
#colSums(is.na(data))
# Total number of rows
total_rows <- nrow(data)

# Count NAs and calculate percentages
na_counts <- colSums(is.na(data))
na_percent <- round((na_counts / total_rows) * 100, 2)

# Create a pretty table
na_table <- data.frame(
  column = names(na_counts),
  na_count = na_counts,
  percent_na = na_percent,
  row.names = NULL
)

# View the full table
print(na_table)
```



SAving my NA table so I can better analyze it 
```{r}
write.csv(na_table, "/Users/ruthermiasadnew/Downloads/na_table.csv", row.names = FALSE)
getwd()
```

here I am dropping all the columns with high NA values.. like above 90% of the data is missing 
also the Hospitalization ICU data which is not relevant for this project. 
```{r}
# Drop columns with >90% missing data
data_cleaned <- data %>%
  select(-c(
    excess_mortality_cumulative_absolute,
    excess_mortality_cumulative,
    excess_mortality,
    excess_mortality_cumulative_per_million,
    icu_patients,
    icu_patients_per_million,
    hosp_patients,
    hosp_patients_per_million,
    weekly_icu_admissions,
    weekly_icu_admissions_per_million,
    weekly_hosp_admissions,
    weekly_hosp_admissions_per_million
  ))
```

Just checking if I actually dropped them 
```{r}
ncol(data)        # Before
ncol(data_cleaned) # After dropping
```

here I am taking a closer look at what each column contains.. like the type and it's unique values.. this will help me better decide how to handle the NA's for each variable 
```{r}
#str(data_cleaned)
for (col in names(data_cleaned)) {
  cat("🔹 Column:", col, "\n")
  cat("📦 Type:", class(data_cleaned[[col]]), "\n")
  cat("🔎 Unique values:\n")
  
  print(unique(data_cleaned[[col]]))  # Shows ALL unique values
  
  cat("\n--------------------------\n")
}


```

Checking again for my cleaned table,, like how many % of the data is N/A
```{r}

# Total number of rows
total_rows <- nrow(data_cleaned)

# Count NAs and calculate percentages
na_counts <- colSums(is.na(data_cleaned))
na_percent <- round((na_counts / total_rows) * 100, 2)

# Create a pretty table
na_table2 <- data.frame(
  column = names(na_counts),
  na_count = na_counts,
  percent_na = na_percent,
  row.names = NULL
)

# View the full table
print(na_table2)
```
Okay here what I tried to do is, positive rate seems like one of the variables that will be very useful for my project.. it tells me the number of positive cases in each country the bad news is it has a lot of missing values like 78% of the data... so my best bet is to fill the missing values by the mean of each country. 

```{r}
library(dplyr)

# Impute missing positive_rate values using each country's mean
data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  mutate(
    positive_rate = ifelse(
      is.na(positive_rate),
      mean(positive_rate, na.rm = TRUE),
      positive_rate
    )
  ) %>%
  ungroup()
```

okay now I will try and squeeze all the relevant data from variables that basically tell me the same thing and those varaibles are... 
total_tests
new_tests
total_tests_per_thousand
new_tests_per_thousand
new_tests_smoothed
new_tests_smoothed_per_thousand
positive_rate
tests_per_case

```{r}
# List of your variables
vars_to_check <- c(
  "total_tests", "new_tests", "total_tests_per_thousand",
  "new_tests_per_thousand", "new_tests_smoothed",
  "new_tests_smoothed_per_thousand", "positive_rate",
  "tests_per_case", "tests_units"
)

# Loop to display unique values for each
for (var in vars_to_check) {
  cat("🔹 Variable:", var, "\n")
  cat("📦 Type:", class(data_cleaned[[var]]), "\n")
  cat("🔎 Unique values:\n")
  print(unique(data_cleaned[[var]]))
  cat("\n------------------------------\n\n")
}


```

so they are all numeric but also the have >80 % missing values so what should we dooo..
  my approach is going to be squezze those variables and get one meaningful variable out of them..so I     can drop them and use my varibale instead. 
  
that variable is the testing effieciency and it will be calculating positive rates and new tests per thousand. so instead of using 8 variables I can use one and get a better result 
```{r}
data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  mutate(
    new_tests_smoothed_per_thousand = ifelse(
      is.na(new_tests_smoothed_per_thousand),
      mean(new_tests_smoothed_per_thousand, na.rm = TRUE),
      new_tests_smoothed_per_thousand
    )
  ) %>%
  ungroup()

```
now I can calculate my testing efficiency variable 
```{r}

sum(is.na(data_cleaned$testing_efficiency))
unique(data$new_tests_smoothed_per_thousand)

dim(data_cleaned)
# Copy the cleaned column from `data` into `data_cleaned`
data_cleaned$new_tests_smoothed_per_thousand <- data$new_tests_smoothed_per_thousand

data_cleaned <- data_cleaned %>%
  mutate(testing_efficiency = scale(new_tests_smoothed_per_thousand)[,1] -
                              scale(positive_rate)[,1])

dim(data_cleaned)
names(data_cleaned)


```
```{r}
library(dplyr)
library(zoo)

data <- data %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(new_tests_smoothed_per_thousand = na.locf(new_tests_smoothed_per_thousand, na.rm = FALSE)) %>%
  ungroup()

data <- data %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(new_tests_smoothed_per_thousand = na.locf(new_tests_smoothed_per_thousand, fromLast = TRUE, na.rm = FALSE)) %>%
  ungroup()

data <- data %>%
  group_by(continent) %>%
  mutate(new_tests_smoothed_per_thousand = ifelse(
    is.na(new_tests_smoothed_per_thousand),
    mean(new_tests_smoothed_per_thousand, na.rm = TRUE),
    new_tests_smoothed_per_thousand
  )) %>%
  ungroup()

data <- data %>%
  mutate(new_tests_smoothed_per_thousand = ifelse(
    is.na(new_tests_smoothed_per_thousand),
    mean(new_tests_smoothed_per_thousand, na.rm = TRUE),
    new_tests_smoothed_per_thousand
  ))
```


```{r}
```


```{r}
```


```{r}
```


making NA its own category for test_units. 
```{r}
# Convert NAs in tests_units into a "Missing" category
data_cleaned <- data_cleaned %>%
  mutate(tests_units = ifelse(is.na(tests_units), "Missing", tests_units)) %>%
  mutate(tests_units = as.factor(tests_units))  # optional: convert to factor
```


```{r}
# Drop testing-related columns after composite feature created
data_cleaned <- data_cleaned %>%
  select(-c(
    total_tests,
    new_tests,
    total_tests_per_thousand,
    new_tests_per_thousand,
    new_tests_smoothed,
    new_tests_smoothed_per_thousand,
    tests_per_case,
  ))
```


```{r}
```


```{r}
```



```{r}
data_cleaned <- data_cleaned %>%
  select(-total_tests, -new_tests, -total_tests_per_thousand, -new_tests_per_thousand,
         -new_tests_smoothed, -new_tests_smoothed_per_thousand,
         -tests_per_case, -tests_units)
```


```{r}
dim(data_cleaned)
names(data_cleaned)
```


```{r}
# Total number of rows
total_rows <- nrow(data_cleaned)

# Count NAs and calculate percentages
na_counts <- colSums(is.na(data_cleaned))
na_percent <- round((na_counts / total_rows) * 100, 2)

# Create a pretty table
na_table3 <- data.frame(
  column = names(na_counts),
  na_count = na_counts,
  percent_na = na_percent,
  row.names = NULL
)

# View the full table
print(na_table3)

```


```{r}
install.packages("pheatmap")  # only if you haven't yet
library(pheatmap)
```


```{r}
vaccine_vars <- c(
  "total_vaccinations",
  "people_vaccinated",
  "people_fully_vaccinated",
  "total_boosters",
  "new_vaccinations",
  "new_vaccinations_smoothed",
  "total_vaccinations_per_hundred",
  "people_vaccinated_per_hundred",
  "people_fully_vaccinated_per_hundred",
  "total_boosters_per_hundred",
  "new_vaccinations_smoothed_per_million",
  "new_people_vaccinated_smoothed",
  "new_people_vaccinated_smoothed_per_hundred"
)
# Check data types
sapply(data_cleaned[vaccine_vars], class)
```


```{r}
# Reuse your cleaned correlation matrix
cor_matrix <- cor(vaccine_vars_clean, use = "pairwise.complete.obs")

# Plot heatmap
pheatmap(cor_matrix,
         color = colorRampPalette(c("red", "white", "blue"))(50),
         fontsize = 9,
         fontsize_row = 9,
         fontsize_col = 9,
         angle_col = 45,
         display_numbers = TRUE,
         main = "💉 Correlation Heatmap of Vaccine Variables")
```


```{r}
dim(data_cleaned)
```
Based on the corrrelation heat map I will be dropping more variables that are highly correlated, because they dont add meaningful insight to my end Goal they are just redudant. 

```{r}
drop_vars <- c(
  "total_vaccinations",
  "people_vaccinated",
  "people_fully_vaccinated",
  "total_boosters",
  "new_vaccinations",
  "new_people_vaccinated_smoothed",
  "new_people_vaccinated_smoothed_per_hundred",
  "people_vaccinated_per_hundred"
)
# data_cleaned <- data_cleaned %>% 
#   select(-all_of(drop_vars))

```

```{r}
dim(data_cleaned)
```
well  stringency_index is very important it tells me how strict the govenmnent was with policy when it came to covid. 
```{r}
data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  mutate(
    stringency_index = ifelse(
      is.na(stringency_index),
      mean(stringency_index, na.rm = TRUE),
      stringency_index
    )
  ) %>%
  ungroup()
```


```{r}
#colSums(is.na(data))
# Total number of rows
total_rows <- nrow(data_cleaned)

# Count NAs and calculate percentages
na_counts <- colSums(is.na(data_cleaned))
na_percent <- round((na_counts / total_rows) * 100, 2)

# Create a pretty table
na_table2 <- data.frame(
  column = names(na_counts),
  na_count = na_counts,
  percent_na = na_percent,
  row.names = NULL
)

# View the full table
print(na_table2)
```
so far this is what we have pookie 

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

# Step 1: Calculate % of missing values for each column
missing_data <- data_cleaned %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "na_count") %>%
  mutate(
    total_rows = nrow(data_cleaned),
    percent_missing = (na_count / total_rows) * 100
  )

# Step 2: Plot missingness as a barplot
ggplot(missing_data, aes(x = reorder(variable, -percent_missing), y = percent_missing)) +
  geom_bar(stat = "identity", fill = "salmon") +
  coord_flip() +
  labs(
    title = "🔍 Percent of Missing Values per Variable",
    x = "Variable",
    y = "Missing Percentage (%)"
  ) +
  theme_minimal(base_size = 10)
```
Boosters - it would make sense that we have a lot of missing data here 
```{r}
#unique(data_cleaned$total_boosters_per_hundred)
data_cleaned$total_boosters_per_hundred[is.na(data_cleaned$total_boosters_per_hundred)] <- 0
sum(is.na(data_cleaned$total_boosters_per_hundred))
```

people_fully_vaccinated_per_hundred- you are getting dropped pookie

also I am imputing 0 for total_vaccinations_per_hundred missing values 
```{r}
# unique(data_cleaned$people_fully_vaccinated_per_hundred)
# cor(
#   data_cleaned$total_vaccinations_per_hundred,
#   data_cleaned$people_fully_vaccinated_per_hundred,
#   use = "complete.obs"
#)#0.9604988 - very high correlation 

data_cleaned <- data_cleaned %>%
  select(-people_fully_vaccinated_per_hundred)

data_cleaned <- data_cleaned %>%
  mutate(total_vaccinations_per_hundred = ifelse(
    is.na(total_vaccinations_per_hundred), 0, total_vaccinations_per_hundred
  ))
sum(is.na(data_cleaned$total_vaccinations_per_hundred))

```
tests_units
```{r}
unique(data_cleaned$tests_units)
data_cleaned <- data_cleaned %>%
  mutate(tests_units = ifelse(is.na(tests_units), "unknown", tests_units))
sum(is.na(data_cleaned$tests_units))

```
```{r}
#colSums(is.na(data))
# Total number of rows
total_rows <- nrow(data_cleaned)

# Count NAs and calculate percentages
na_counts <- colSums(is.na(data_cleaned))
na_percent <- round((na_counts / total_rows) * 100, 2)

# Create a pretty table
na_table3 <- data.frame(
  column = names(na_counts),
  na_count = na_counts,
  percent_na = na_percent,
  row.names = NULL
)

# View the full table
print(na_table3)

```

handwashing_facilities- 
```{r}
unique(data_cleaned$handwashing_facilities)
data_cleaned <- data_cleaned %>%
  mutate(
    handwashing_missing = ifelse(is.na(handwashing_facilities), 1, 0),
    handwashing_facilities = ifelse(is.na(handwashing_facilities), 0, handwashing_facilities)
  )
sum(is.na(data_cleaned$handwashing_facilities))
```

	
reproduction_rate
```{r}
#unique(data_cleaned$reproduction_rate)

#by country pookie
data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  mutate(reproduction_rate = ifelse(
    is.na(reproduction_rate),
    mean(reproduction_rate, na.rm = TRUE),
    reproduction_rate
  )) %>%
  ungroup()
sum(is.na(data_cleaned$reproduction_rate))

# Add a missingness flag for reproduction_rate
library(dplyr)

data_cleaned <- data_cleaned %>%
  mutate(reproduction_rate_missing = ifelse(is.na(reproduction_rate), 1, 0))

```

new_vaccinations_smoothed and new_vaccinations_smoothed_per_million tell me the exact same thing 

```{r}
unique(data_cleaned$new_vaccinations_smoothed_per_million)
# Drop the raw version
data_cleaned <- data_cleaned %>%
  select(-new_vaccinations_smoothed)

# Impute 0 for per million version
data_cleaned <- data_cleaned %>%
  mutate(new_vaccinations_smoothed_per_million = ifelse(
    is.na(new_vaccinations_smoothed_per_million),
    0,
    new_vaccinations_smoothed_per_million
  ))

sum(is.na(data_cleaned$new_vaccinations_smoothed_per_million))

```

extreme_poverty - 
```{r}
unique(data_cleaned$extreme_poverty)
data_cleaned <- data_cleaned %>%
  mutate(extreme_poverty = ifelse(is.na(extreme_poverty), 0, extreme_poverty))
sum(is.na(data_cleaned$extreme_poverty))
```
male and female smokers.. well I want to join them cause I am not working with gender here.. 
```{r}
data_cleaned <- data_cleaned %>%
  mutate(avg_smokers = rowMeans(select(., female_smokers, male_smokers), na.rm = TRUE))
data_cleaned <- data_cleaned %>%
  select(-female_smokers, -male_smokers)
data_cleaned$avg_smokers[is.na(data_cleaned$avg_smokers)] <- 0
sum(is.na(data_cleaned$avg_smokers))
unique(data_cleaned$avg_smokers)
```

hospital_beds_per_thousand
```{r}
# data_cleaned <- data_cleaned %>%
#   group_by(location) %>%
#   mutate(hospital_beds_per_thousand = ifelse(
#     is.na(hospital_beds_per_thousand),
#     mean(hospital_beds_per_thousand, na.rm = TRUE),
#     hospital_beds_per_thousand
#   )) %>%
#   ungroup()
# sum(is.na(data_cleaned$hospital_beds_per_thousand))
# unique(data_cleaned$hospital_beds_per_thousand)

# Load necessary library
library(dplyr)

# Select relevant columns and remove rows with NA
cor_data <- data_cleaned %>%
  select(hospital_beds_per_thousand, reproduction_rate, positive_rate, new_cases_per_million) %>%
  na.omit()

# Compute the correlation matrix
cor_matrix <- cor(cor_data)

# Print it nicely
print(round(cor_matrix, 3))

#hospital beds has a low correlation with the potentially key variables so I will drop it 
# Drop hospital_beds_per_thousand
data_cleaned <- data_cleaned %>%
  select(-hospital_beds_per_thousand)
names(data_cleaned)
```

positive_rates 
```{r}
unique(data_cleaned$positive_rate)
     data_cleaned <- data_cleaned %>%
  mutate(
    positive_rate_missing = ifelse(is.na(positive_rate), 1, 0),
    positive_rate = ifelse(is.na(positive_rate), 0, positive_rate)
  )

sum(is.na(data_cleaned$positive_rate))
names(data_cleaned)
```

testing_efficiency

```{r}
sum(is.na(data_cleaned$testing_efficiency))

unique(data_cleaned$positive_rate)

```

human_development_index
```{r}
sum(is.na(data_cleaned$human_development_index))
data_cleaned <- data_cleaned %>%
  group_by(continent) %>%
  mutate(human_development_index = ifelse(
    is.na(human_development_index),
    mean(human_development_index, na.rm = TRUE),
    human_development_index
  )) %>%
  ungroup()

unique(data_cleaned$human_development_index)

```

cardiovasc_death_rate
```{r}
sum(is.na(data_cleaned$cardiovasc_death_rate))


data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(cardiovasc_death_rate = na.locf(cardiovasc_death_rate, na.rm = FALSE)) %>%  # forward fill
  mutate(cardiovasc_death_rate = na.locf(cardiovasc_death_rate, fromLast = TRUE, na.rm = FALSE)) %>%  # backward fill
  ungroup()

data_cleaned <- data_cleaned %>%
  group_by(continent) %>%
  mutate(cardiovasc_death_rate = ifelse(
    is.na(cardiovasc_death_rate),
    mean(cardiovasc_death_rate, na.rm = TRUE),
    cardiovasc_death_rate
  )) %>%
  ungroup()

unique(data_cleaned$cardiovasc_death_rate)
```


stringency_index
```{r}
sum(is.na(data_cleaned$stringency_index))


data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(stringency_index = na.locf(stringency_index, na.rm = FALSE)) %>%  # Forward fill
  mutate(stringency_index = na.locf(stringency_index, fromLast = TRUE, na.rm = FALSE)) %>%  # Backward fill
  ungroup()

data_cleaned <- data_cleaned %>%
  group_by(continent) %>%
  mutate(stringency_index = ifelse(
    is.na(stringency_index),
    mean(stringency_index, na.rm = TRUE),
    stringency_index
  )) %>%
  ungroup()
unique(data_cleaned$stringency_index)

data_cleaned <- data_cleaned %>%
  mutate(stringency_index = ifelse(
    is.na(stringency_index),
    mean(stringency_index, na.rm = TRUE),
    stringency_index
  ))
```


population_density
```{r}
sum(is.na(data_cleaned$population_density))
# Forward fill + backward fill by country (works if some rows in the same country have values)
library(zoo)
data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(population_density = na.locf(population_density, na.rm = FALSE)) %>%
  mutate(population_density = na.locf(population_density, fromLast = TRUE, na.rm = FALSE)) %>%
  ungroup()

# If there are still NAs, use continent mean
data_cleaned <- data_cleaned %>%
  group_by(continent) %>%
  mutate(population_density = ifelse(
    is.na(population_density),
    mean(population_density, na.rm = TRUE),
    population_density
  )) %>%
  ungroup()


unique(data_cleaned$population_density)
```




median_age
```{r}
sum(is.na(data_cleaned$median_age))

data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(median_age = na.locf(median_age, na.rm = FALSE)) %>%
  mutate(median_age = na.locf(median_age, fromLast = TRUE, na.rm = FALSE)) %>%
  ungroup()

data_cleaned <- data_cleaned %>%
  group_by(continent) %>%
  mutate(median_age = ifelse(
    is.na(median_age),
    mean(median_age, na.rm = TRUE),
    median_age
  )) %>%
  ungroup()

unique(data_cleaned$continent)
```


aged_65_older
```{r}
sum(is.na(data_cleaned$aged_65_older))

data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(aged_65_older = na.locf(aged_65_older, na.rm = FALSE)) %>%
  mutate(aged_65_older = na.locf(aged_65_older, fromLast = TRUE, na.rm = FALSE)) %>%
  ungroup()

data_cleaned <- data_cleaned %>%
  group_by(continent) %>%
  mutate(aged_65_older = ifelse(
    is.na(aged_65_older),
    mean(aged_65_older, na.rm = TRUE),
    aged_65_older
  )) %>%
  ungroup()

unique(data_cleaned$aged_65_older)

```


aged_70_older
```{r}
sum(is.na(data_cleaned$aged_70_older))


```


gdp_per_capita

```{r}
sum(is.na(data_cleaned$gdp_per_capita))


```

reproduction_rate
```{r}
sum(is.na(data_cleaned$reproduction_rate))

```


```{r}
library(dplyr)
library(zoo)

columns_to_clean <- c("aged_70_older", "gdp_per_capita", "reproduction_rate")

for (col in columns_to_clean) {
  data_cleaned <- data_cleaned %>%
    group_by(location) %>%
    arrange(date) %>%
    mutate(!!sym(col) := na.locf(!!sym(col), na.rm = FALSE)) %>%
    mutate(!!sym(col) := na.locf(!!sym(col), fromLast = TRUE, na.rm = FALSE)) %>%
    ungroup()
  
  data_cleaned <- data_cleaned %>%
    group_by(continent) %>%
    mutate(!!sym(col) := ifelse(
      is.na(!!sym(col)),
      mean(!!sym(col), na.rm = TRUE),
      !!sym(col)
    )) %>%
    ungroup()
}
```


```{r}
data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(diabetes_prevalence = na.locf(diabetes_prevalence, na.rm = FALSE)) %>%
  mutate(diabetes_prevalence = na.locf(diabetes_prevalence, fromLast = TRUE, na.rm = FALSE)) %>%
  ungroup() %>%
  group_by(continent) %>%
  mutate(diabetes_prevalence = ifelse(
    is.na(diabetes_prevalence),
    mean(diabetes_prevalence, na.rm = TRUE),
    diabetes_prevalence
  )) %>%
  ungroup() %>%
  mutate(diabetes_prevalence = ifelse(
    is.na(diabetes_prevalence),
    mean(diabetes_prevalence, na.rm = TRUE),
    diabetes_prevalence
  ))
```

```{r}
data_cleaned <- data_cleaned %>%
  group_by(location) %>%
  arrange(date) %>%
  mutate(life_expectancy = na.locf(life_expectancy, na.rm = FALSE)) %>%
  mutate(life_expectancy = na.locf(life_expectancy, fromLast = TRUE, na.rm = FALSE)) %>%
  ungroup() %>%
  group_by(continent) %>%
  mutate(life_expectancy = ifelse(
    is.na(life_expectancy),
    mean(life_expectancy, na.rm = TRUE),
    life_expectancy
  )) %>%
  ungroup()
```


```{r}
sum(is.na(data_cleaned$continent))
```


```{r}
library(dplyr)
library(zoo)

cols_total <- c("total_cases", "total_deaths", "total_cases_per_million", "total_deaths_per_million")

for (col in cols_total) {
  data_cleaned <- data_cleaned %>%
    group_by(location) %>%
    arrange(date) %>%
    mutate(!!sym(col) := na.locf(!!sym(col), na.rm = FALSE)) %>%
    ungroup()
}
```


```{r}
cols_new <- c("new_cases", "new_deaths", "new_cases_per_million", "new_deaths_per_million")

for (col in cols_new) {
  data_cleaned[[col]][is.na(data_cleaned[[col]])] <- 0
}
```


```{r}
#data_cleaned <- data_cleaned %>% filter(!is.na(continent))

library(dplyr)
library(zoo)

# 1. Define columns to fix
cols_total <- c("total_cases", "total_deaths", 
                "total_cases_per_million", "total_deaths_per_million")
cols_smoothed <- c("new_cases_smoothed", "new_deaths_smoothed", 
                   "new_cases_smoothed_per_million", "new_deaths_smoothed_per_million")

# 2. Forward fill & backward fill total and smoothed columns
for (col in c(cols_total, cols_smoothed)) {
  data_cleaned <- data_cleaned %>%
    group_by(location) %>%
    arrange(date) %>%
    mutate(!!sym(col) := na.locf(!!sym(col), na.rm = FALSE)) %>%
    mutate(!!sym(col) := na.locf(!!sym(col), fromLast = TRUE, na.rm = FALSE)) %>%
    ungroup()
}

# 3. Replace any final NAs in totals/smoothed with 0
for (col in c(cols_total, cols_smoothed)) {
  data_cleaned[[col]][is.na(data_cleaned[[col]])] <- 0
}
```

```{r}
write.csv(data_cleaned, "data_cleaned.csv", row.names = FALSE)

```

```{r}
unique(data$location)
```


```{r}
c("England", "Wales", "Taiwan", "Scotland", "Northern Ireland", "Macao") %in% unique(data$location)
```


```{r}
library(dplyr)

data %>%
  filter(location %in% c("England", "Wales", "Taiwan", "Scotland", "Northern Ireland", "Macao", "North Korea")) %>%
  group_by(location) %>%
  summarise(row_count = n())
```


```{r}
data %>%
  filter(location %in% c("England", "Wales", "Taiwan", "Scotland", "Northern Ireland", "Macao", "North Korea")) %>%
  group_by(location) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE)) %>%
  arrange(total_cases)
```


```{r}
data %>%
  filter(location %in% c("England", "Wales", "Taiwan", "Scotland", "Northern Ireland", "Macao", "North Korea")) %>%
  group_by(location) %>%
  summarise(
    total_NAs = sum(is.na(total_cases)),
    total_nonNAs = sum(!is.na(total_cases)),
    total_cases_sum = sum(total_cases, na.rm = TRUE)
  )
```


```{r}
unique(data$location)[str_detect(unique(data$location), "United|UK|Kingdom")]
```


```{r}
# Install (if needed) and load packages
install.packages("naniar")      # For missingness heatmap
install.packages("ggplot2")     # For plotting
library(naniar)
library(ggplot2)

# Assuming your cleaned dataset is named `data_cleaned`

# 🔥 Heatmap of Missing Values
gg_miss_var(data, show_pct = TRUE) +
  labs(
    title = "🔥 Missing Data  by Variable",
    x = "Variable",
    y = "Number of Missing Values"
  ) +
  theme_minimal(base_size = 8)
```


```{r}
# Install and load libraries (run only once)

# Make sure your cleaned data is called data_cleaned

# 🧯Heatmap-style visualization (tile map) of missing values
vis_miss(data, sort_miss = TRUE, cluster = TRUE) +
  labs(
    title = "🧊 Heatmap of Missingness in COVID-19 Variables",
    subtitle = "Each tile shows whether the value is missing (NA) or observed",
    x = "Variables",
    y = "Rows / Observations"
  ) +
  theme_minimal(base_size = 8) +  # smaller font
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  )
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

